Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-18 analysis/paper/paper.Rmd
worked on the background section

<!--Info about Yanyuan area-->
The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)). Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.

<!--Info about Guijiabao-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the Basin is the only site with continuous occupation from the early Neolithic period to the historical period. The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022]. The Neolithic period can be further separated into two subphases, 5000-4500 cal. BP and 4500-3700 cal. BP, according to the cultural remains and pottery characteristics. Recent archaeological evidence indicates the spread of millet-rice mixed economy to southwest Sichuan can be dated back to around 5,000 BP.

@@

## Methods
In this study, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial or marine animals and plants, have specific ranges of δ^13^C values [@Evershed1994].

>>>>

<!--Info about Yanyuan area-->
The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)). Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.

<!--Info about Guijiabao-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is an unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and remains, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long term-occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal. BP and 4500-3700 cal. BP, according to the artifact assemblages and ceramic forms and styles. The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and a mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022]

@@

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial or marine animals and plants, have specific ranges of δ^13^C values [@Evershed1994].
